$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-04 paper/paper.tex
Discard unnecessary lemma, group auxiliary results

    \item \textbf{Leftover typing}: The typing judgments use \emph{leftover typing} to model the resource-aware \picalc{} (in \autoref{leftover-typing}).
    This approach adds a leftover usage context to the typing judgments.
    Typing derivations take the resources of their input usage context, consume some of them, and leave the rest as leftovers in the output usage context.
    One of the benefits of this technique is that it removes the need for extrinsic context splits, which are rendered unnecessary.
  \end{itemize}
  \item \textbf{Machine verified proofs of subject reduction and auxiliary theorems}:
  \autoref{type-safety} details the type safety properties of our \picalc{} with leftovers.
  Leftover typing allows \emph{framing} (\autoref{framing}), \emph{weakening} (\autoref{weakening}) and \emph{strengthening} (\autoref{strengthening}) theorems to be stated.
  We use these to prove \emph{subject congruence} (\autoref{subject-congruence}), and together with \emph{substitution} (\autoref{substitution}), to prove \emph{subject reduction} (\autoref{subject-reduction}).

@@

We use \emph{leftover typing} \cite{Allais2018a} for our type system, an approach that, in addition to the usual typing context and usage context, adds an extra output usage context to the typing rules.
This output context contains the \emph{leftovers} (the unused multiplicities) of the process being typed.
These leftovers can then be used as an input to another typing derivation.
Leftover typing presents three main advantages:
\begin{itemize}
  \item
    The transformations to the usage context (consequence of receiving and sending over a channel) are encapsulated into a variable reference judgment with both an input and an output usage context.
  \item
    No need for an extensional context splitting proof for parallel composition.
    Such a proof would need to construct a witness $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$, where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    This could be obtained through automated search --- but it is better yet not to need it.
    Leftover typing uses the leftovers of $P$ to type $Q$.
  \item
    New theorems like \textbf{framing} (\autoref{framing}), \textbf{weakening} (\autoref{weakening}) and \textbf{strengthening} (\autoref{strengthening}) can be stated thanks to the presence of a leftover context.
\end{itemize}

@@

\todo{intro}
\todo{I wonder if we can remove the proofs everywhere and comment on them in the intro and where they can be found.}
\subsection{Framing}
\label{framing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Framing captures the idea that the well-typedness of a process is independent of its leftover resources.
\todo{Do not forward refer when it comes to results. Never refer to results in the appendix. Consider that, if the paper get publshed, it will get so without the appendix. It should be *completely* self contained}
\begin{nitheorem}
  \label{thm:framing}
  Let $P$ be a well-typed process $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$.
  Then, per \autoref{lm:types-op}, there exists some $\Delta$ such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Take arbitrary contexts $\Gamma_r$ and $\Xi_r$ such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\subsection{Weakening}
\label{weakening}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Inserting a new variable into the context preserves the well-typedness of a process as long as the usage annotation of the inserted variable is preserved as a leftover.

\begin{nitheorem}
  \label{thm:weakening}
  Let $\func{insert}_i$ insert an element into a context at position $i$ --- for simplicity, we use it both to insert types into typing contexts and usage annotations into usage contexts.
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then lifting every variable greater than or equal to $i$ in $P$ is well-typed under $\types{\func{insert}_i \; t \; \gamma}{\func{insert}_i \; x \; \Gamma}{\func{lift}_i \; P}{\func{insert}_i \; x \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\subsection{Strengthening}
\label{strengthening}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Removing an unused variable preserves the well-typedness of a process.
We say a variable is \emph{unused} within a process if neither inputs nor outputs refer to it.

@@

\begin{nitheorem}
  \label{thm:strengthening}
  Let $\func{delete}_i$ delete the element at position $i$ from a context --- for simplicity, we use it both to delete types from typing contexts and usage annotations from usage contexts.
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable index not mentioned by $P$, such that $\Unused_i \; P$.
  Then lowering every variable greater than or equal to $i$ in $P$ is well-typed under $\types{\func{delete}_i \; \gamma}{\func{delete}_i \; \Gamma}{\func{lower}_i \; P}{\func{delete}_i \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\subsection{Swapping}
\label{swapping}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Swapping two variables preserves the well-typedness of a process.

@@

\begin{nitheorem}
  \label{thm:swapping}
  Let $\func{swap}_i$ be a function that swaps variables $i$ and $\suc i$.
  For simplicity, we overload $\func{swap}$ and use it to swap types in a typing context, multiplicities in a usage context, and variable references in a process.
  \todo{}
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then $\types{\func{swap}_i \; \gamma}{\func{swap}_i \; \Gamma}{\func{swap}_i \; P}{\func{swap}_i \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\todo{Maybe all previous subsections could be put into one "Auxiliary Results"}
\todo{Replace the proofs with an intuition on the theorem}
\subsection{Subject Congruence}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{subject-congruence}

@@

\todo{maybe only focus on the first case, internal and external and say the rest is by induction blah bla and details can be found in appenix/or github}
\begin{proof}
  \hfill{}\\
  \begin{itemize}
    \item
    In the $\constr{comm}$ case we apply framing (\autoref{thm:framing}) to rearrange the assumptions and then apply substitution (\autoref{thm:substitution}) and strengthening (\autoref{thm:strengthening}).

>>>>

    \item \textbf{Leftover typing}: The typing judgments use \emph{leftover typing} to model the resource-aware \picalc{} (in \autoref{leftover-typing}).
    This approach adds a leftover usage context to the typing judgments.
    Typing derivations take the resources of their input usage context, consume some of them, and leave the rest as leftovers in the output usage context.
    One of the benefits of this technique is that it removes the need for extrinsic context splits, which are rendered unnecessary.
  \end{itemize}
  \item \textbf{Machine verified proofs of subject reduction and auxiliary theorems}:
  \autoref{type-safety} details the type safety properties of our \picalc{} with leftovers.
  Leftover typing allows \emph{framing} (\autoref{thm:framing}), \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}) theorems to be stated.
  We use these to prove \emph{subject congruence} (\autoref{subject-congruence}), and together with \emph{substitution} (\autoref{substitution}), to prove \emph{subject reduction} (\autoref{subject-reduction}).

@@

We use \emph{leftover typing} \cite{Allais2018a} for our type system, an approach that, in addition to the usual typing context and usage context, adds an extra output usage context to the typing rules.
This output context contains the \emph{leftovers} (the unused multiplicities) of the process being typed.
These leftovers can then be used as an input to another typing derivation.
Leftover typing presents three main advantages:
\begin{itemize}
  \item
    The transformations to the usage context (consequence of receiving and sending over a channel) are encapsulated into a variable reference judgment with both an input and an output usage context.
  \item
    No need for an extensional context splitting proof for parallel composition.
    Such a proof would need to construct a witness $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$, where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    This could be obtained through automated search --- but it is better yet not to need it.
    Leftover typing uses the leftovers of $P$ to type $Q$.
  \item
    New theorems like \textbf{framing} (\autoref{thm:framing}), \textbf{weakening} (\autoref{thm:weakening}) and \textbf{strengthening} (\autoref{thm:strengthening}) can be stated thanks to the presence of a leftover context.
\end{itemize}

@@

\todo{intro}
\todo{I wonder if we can remove the proofs everywhere and comment on them in the intro and where they can be found.}

\subsection{Auxiliaty Results}
\label{auxiliary-results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{nitheorem}
  \label{thm:framing}
  Framing captures the idea that the well-typedness of a process is independent of its leftover resources.
  Let $P$ be a well-typed process $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$.
  Let there be a $\Delta$ such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Take arbitrary contexts $\Gamma_r$ and $\Xi_r$ such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{nitheorem}
  \label{thm:weakening}
  Inserting a new variable into the context preserves the well-typedness of a process as long as the usage annotation of the inserted variable is preserved as a leftover.

  Let $\func{insert}_i$ insert an element into a context at position $i$ --- for simplicity, we use it both to insert types into typing contexts and usage annotations into usage contexts.
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then lifting every variable greater than or equal to $i$ in $P$ is well-typed under $\types{\func{insert}_i \; t \; \gamma}{\func{insert}_i \; x \; \Gamma}{\func{lift}_i \; P}{\func{insert}_i \; x \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{nitheorem}
  \label{thm:strengthening}
  Removing an unused variable preserves the well-typedness of a process.

  Let $\func{delete}_i$ delete the element at position $i$ from a context --- for simplicity, we use it both to delete types from typing contexts and usage annotations from usage contexts.
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable index not mentioned by $P$, such that $\Unused_i \; P$.
  Then lowering every variable greater than or equal to $i$ in $P$ is well-typed under $\types{\func{delete}_i \; \gamma}{\func{delete}_i \; \Gamma}{\func{lower}_i \; P}{\func{delete}_i \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{nitheorem}
  \label{thm:swapping}
  Swapping two variables preserves the well-typedness of a process.
  Let $\func{swap}_i$ be a function that swaps variables $i$ and $\suc i$.
  For simplicity, we overload $\func{swap}$ and use it to swap types in a typing context, multiplicities in a usage context, and variable references in a process.
  \todo{}
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then $\types{\func{swap}_i \; \gamma}{\func{swap}_i \; \Gamma}{\func{swap}_i \; P}{\func{swap}_i \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\todo{Replace the proofs with an intuition on the theorem}
\subsection{Subject Congruence}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{subject-congruence}

@@

\begin{proof}
  \hfill{}\\
  \begin{itemize}
    \item
    In the $\constr{comm}$ case we apply framing (\autoref{thm:framing}) to rearrange the assumptions and then apply substitution (\autoref{thm:substitution}) and strengthening (\autoref{thm:strengthening}).
